---
title: "Vulnerability to Covid-19 in Scotland"
subtitle: "Analysis by Intermediate Zone"
author: "David Henderson"
date: "30/03/2020"
output: 
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    theme: spacelab
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

Recreate analysis in `prescribing.Rmd` but use Intermediate zones (as opposed to Health Board) as the geogrpahy. 


## Packages

Load in required packages

```{r, message=FALSE,warning=FALSE}
library(tidyverse)
library(ggrepel)
library(janitor)
library(gtools)
library(sf)
library(cowplot)

theme_set(theme_minimal(base_family = "Roboto",
                        base_size = 18) +
            theme(panel.grid.minor = element_blank()))
```

## Data

Measurement data comes from [NHS Scotland's open data pages](https://www.opendata.nhs.scot/theme/isd). I use the most recent prescribing data available from [Dec 2019](https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community/resource/fa276ad2-669a-472f-9c47-809f199fae21). The geography lookup codes were derived from the [datazone 2011 file](https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/395476ab-0720-4740-be07-ff4467141352) (although in hindsight I could have just as easily used the [Health Board lookup](https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc)]). Population estimates for each health board were derived from [this file](https://www.opendata.nhs.scot/dataset/population-estimates/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1) using the most recent estimates from 2018. 

Some of these files only need some very quick cleaning to either drop unnecessary information or quick calculations so do those in this chunk of code where appropriate. 

*Note:* the first of these files is very large and can take a few minutes to load. 

```{r, warning=FALSE, message=FALSE}
#Load in prescribing data direct from source. This is a very large file 
#so using data.table's fread() function

presc <- data.table::fread("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/fa276ad2-669a-472f-9c47-809f199fae21/download/pitc201912.csv")

#Load in Health Board lookup file to match HB codes to names
#Some wuick tidying here to drop unnecessary data in this file
geog <- data.table::fread("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/395476ab-0720-4740-be07-ff4467141352/download/geography_codes_and_labels_dz2011_10012020.csv") %>% 
  clean_names %>% 
  select(dz2011, iz2011, iz2011name) 


dz_pop <- data.table::fread("https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/c505f490-c201-44bd-abd1-1bd7a64285ee/download/dz2011-pop-est_30082019.csv") %>%  
  mutate(Year = as.character(Year)) %>% 
  filter(Year == "2018" & DZ2011 != "S92000003") %>% 
  clean_names

gp <- data.table::fread("https://www.opendata.nhs.scot/dataset/f23655c3-6e23-4103-a511-a80d998adb90/resource/a444ae58-9f8c-4447-90d8-5c8641171bb7/download/practice_contactdetails_jan2020-open-data.csv") %>% 
  clean_names %>% 
  select(practice_code, practice_list_size, nhs_board, data_zone2011 ) %>% 
  rename(dz2011 = data_zone2011)
```


```{r}
dz_pop %>% 
  #select rows from 70 plus only
  select(dz2011, age70:age90plus) %>%
  #calculate total population over 70 in a new column
  mutate(pop_70plus = rowSums(.[2:22])) %>% 
  #drop all the other age columns
  select(dz2011, pop_70plus) %>% 
  left_join(., geog) %>% 
  group_by(iz2011) %>% 
  summarise(pop_70plus = sum(pop_70plus)) -> iz_pop
geog
```


```{r}
presc_resp <- 
  presc %>% 
  #names to lower case
  clean_names %>% 
  #tidy up the HB name
  rename(hb2014 = hbt2014, 
         practice_code = gp_practice) %>% 
  #Keep only rows from chapter 3
  filter(str_detect(bnf_item_code, "^03")) %>% 
  left_join(., gp) %>% 
  left_join(., geog) %>% 
  select(iz2011, iz2011name, paid_quantity) %>% 
  group_by(iz2011, iz2011name) %>% 
  summarise(paid_quantity = sum(paid_quantity))

presc_resp
```



```{r}
dz_pop %>% 
  #select rows from 70 plus only
  select(dz2011, age70:age90plus) %>%
  #calculate total population over 70 in a new column
  mutate(pop_70plus = rowSums(.[2:22])) %>% 
  #drop all the other age columns
  select(dz2011, pop_70plus) %>% 
  left_join(., geog) %>% 
  group_by(iz2011) %>% 
  summarise(pop_70plus = sum(pop_70plus)) -> iz_pop
iz_pop

dz_pop %>% 
  select(dz2011, all_ages) %>% 
  group_by(dz2011) %>% 
  summarise(all_ages = sum(all_ages)) %>% 
  left_join(geog) %>% 
  group_by(iz2011) %>% 
  summarise(all_ages = sum(all_ages)) %>% 
  left_join(., iz_pop) %>% 
  full_join(presc_resp, .) %>% 
  select(iz2011, iz2011name, all_ages, pop_70plus, paid_quantity) -> final
```


```{r}
final
```

```{r}
dz_pop %>% 
  group_by(dz2011) %>% 
  summarise(dz_pop = sum(all_ages)) %>% 
  left_join(., gp) %>% 
  left_join(presc_resp, .) %>% 
  left_join(., geog) %>%
  left_join(., iz_pop) %>% 
  select(iz2011, dz2011, practice_code, number_of_paid_items,
         paid_quantity, dz_pop, practice_list_size, pop_70plus) %>% 
  group_by(iz2011) %>% 
  mutate(pop_70plus = round(pop_70plus/sum(dz_pop) *100, 1),
         resp = sum(paid_quantity)/sum(dz_pop)) %>% 
  distinct(iz2011, pop_70plus, resp)
```


```{r}
dz_pop %>% 
  #select rows from 70 plus only
  select(dz2011, age70:age90plus) %>%
  #calculate total population over 70 in a new column
  mutate(pop_70plus = rowSums(.[2:22])) %>% 
  #drop all the other age columns
  select(dz2011, pop_70plus) %>% 
  left_join(., geog) %>% 
  group_by(iz2011) %>% 
  summarise(pop_70plus = sum(pop_70plus)) -> iz_pop
```



