---
title: "Lanarkshire"
subtitle: "Analysis by NHS Health Board"
author: "David Henderson"
date: "30/03/2020"
output: 
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Code to accompany blog post on personal and SCADR pages. 

## Packages

Load in required packages

```{r, message=FALSE,warning=FALSE}
library(tidyverse)
library(ggrepel)
library(ggthemes)
library(janitor)
library(gtools)
library(sf)
library(cowplot)

theme_set(theme_minimal(base_family = "Roboto",
                        base_size = 18) +
            theme(panel.grid.minor = element_blank(),
                  plot.margin = margin(3,15,3,3,"mm")))
options(scipen = 10)
```

## Data

Measurement data comes from [NHS Scotland's open data pages](https://www.opendata.nhs.scot/theme/isd). I use the most recent prescribing data available from [Dec 2019](https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community/resource/fa276ad2-669a-472f-9c47-809f199fae21). The geography lookup codes were derived from the [datazone 2011 file](https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/395476ab-0720-4740-be07-ff4467141352) (although in hindsight I could have just as easily used the [Health Board lookup](https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc)]). Population estimates for each health board were derived from [this file](https://www.opendata.nhs.scot/dataset/population-estimates/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1) using the most recent estimates from 2018. 

Some of these files only need some very quick cleaning to either drop unnecessary information or quick calculations so do those in this chunk of code where appropriate. 

*Note:* the first of these files is very large and can take a few minutes to load. 

```{r, warning=FALSE, message=FALSE}
#Load in prescribing data direct from source. This is a very large file 
#so using data.table's fread() function

presc <- data.table::fread("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/fa276ad2-669a-472f-9c47-809f199fae21/download/pitc201912.csv")

#Load in Health Board lookup file to match HB codes to names
#Some wuick tidying here to drop unnecessary data in this file
geog <- data.table::fread("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/395476ab-0720-4740-be07-ff4467141352/download/geography_codes_and_labels_dz2011_10012020.csv") %>% 
  clean_names %>% 
  select(hb2014, hb2014name) %>% 
  distinct(hb2014, hb2014name)

#Load in popuoation estimates
#More quick cleaning - keep most recent year of data and drop data 
#For Scotland as a whole
pop <- data.table::fread("https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1/download/hb2014_pop_est_01072019.csv") %>% 
  mutate(Year = as.character(Year)) %>% 
  filter(Year == "2018" & HB2014 != "S92000003") 

#Quickly calculate population for each health board. 
hb_pop <- 
  pop %>% 
  select(HB2014, AllAges) %>% 
  group_by(HB2014) %>% 
  summarise(total_pop = sum(AllAges)) %>% 
  clean_names
```

# Tidy the prescribing data

Here what the prescribing file looks like

**Code below to create plot showing NHS Lanakshire as an outlier**

```{r, fig.width=12}
presc %>% 
  rename(hb2014 = HBT) %>% 
  left_join(., geog) %>% 
  left_join(., hb_pop) %>% 
  group_by(hb2014name) %>% 
  summarise(n = n(),
            pop = max(total_pop)) %>% 
  ungroup -> lanark

lanark$lab <- lanark$hb2014name %in% "NHS Lanarkshire"

lanark %>% 
  ggplot(aes(pop, n, colour = lab)) +
  geom_point() +
  geom_text_repel(data = subset(lanark, lab),
                  aes(label = hb2014name)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_ptol(guide = guide_none()) +
  expand_limits(x = c(0, 1200000), y = c(0, 300000)) +
  coord_cartesian(clip = "off") +
  labs(title = "Community prescribing data",
       subtitle = "Dec 2019 file", 
       x = "Health Board Population",
       y = "Number of rows in file") -> lanark_plot
lanark_plot
```



